---
title: "Analysis 1-5"
author: "Michèle, Lennart, Michel, Sebastian"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
library("tidyverse")
library("uwot")
library("factoextra")
```





# PART 1



## load data (data given by advisor)

The data frame is loaded as describe on the Github page for project 3.
```{r}
MS_Table <- read.table("RDeeP_HeLa_Mitosis.csv", header = TRUE, row.names = 1, sep = ";")
```

```{r}
head(rownames(MS_Table),12)
```

```{r}
head(colnames(MS_Table),12)
```

## data cleanup

First, we want to check whether every column's data type is numeric.
```{r}
sum(apply(MS_Table, 2, is.numeric)) == ncol(MS_Table)
glimpse(MS_Table)
```

Every column's data type is numeric (double)! 

We want to remove every row that has only zeros or negative numbers in at least one of the 3 replicates for each condition (RNase or Ctrl). Those rows will not help us to identify shifts and therefore to identify RNA-depending proteins.
```{r}
# find all proteins that contain only zeros in at least one replicate

del_row <- c()
# save those proteins (row number) in del_row

for (i in c("Ctrl_Rep1", "Ctrl_Rep2", "Ctrl_Rep3", "RNase_Rep1", "RNase_Rep2", "RNase_Rep3")) {
  # iterate through all replicates
  
  for (j in 1:nrow(MS_Table)) {
    # iterate through all values
    
    if (sum(MS_Table[j, grep(i, colnames(MS_Table))]) <= 0) {
      del_row <- append(del_row, j)
    }
  }
}
del_row_sort <- sort(unique(del_row))
del_row_sort
```

Therefore, we remove rows 691, 3045, 3575, 4057, 5568, 6732 and 7112.
```{r}
# remove the rows saved in del_row_sort
MS_Table <- MS_Table[-del_row_sort,]
```

All in all, we removed 7 rows that contained only zeros in at least one of the replicates of the two conditions (RNase or Control).

## describing the structure of our data

Dimension of our data frame:
```{r}
dim(MS_Table)
```

Some row names:
```{r}
head(rownames(MS_Table))
```

Some column names:
```{r}
head(colnames(MS_Table))
```

## normalize replicates

In the first normalization step, we want to have a look at the replicates of every fraction under every condition. Theoretically, the three replicates of each fraction are equal, due to them being the results of the same experiment. However, due to experimental differences, this usually is not the case. Therefore, we normalize the replicates in order to make them comparable. To do so, we normalize every three replicates of a given fraction and condition by computing their sums and then computing the mean value of the two closest sums. The quotients of this mean sum and the three sums of the replicates will give us the normalization factors for the three replicates.

```{r}
for (i in seq(1, 148, by = 3)) {
# iterate through the fractions (each fraction has 3 replicates)  
  
  #print(i)
  
  # compute the sums of every replicate
  sums <- c(sum(MS_Table[, i]), sum(MS_Table[, i + 1]), sum(MS_Table[, i + 2]))
  
  # compute the mean value of the two closest sums
  if (abs(sums[1] - sums[2]) < abs(sums[1] - sums[3]) &
      abs(sums[1] - sums[2]) < abs(sums[2] - sums[3])) {
    mean_sum <- mean(c(sums[1], sums[2]))
  } else if (abs(sums[2] - sums[3]) < abs(sums[1] - sums[2]) &
             abs(sums[2] - sums[3]) < abs(sums[1] - sums[3])) {
    mean_sum <- mean(c(sums[2], sums[3]))
  } else {
    mean_sum <- mean(c(sums[1], sums[3]))
  }
  #print(mean_sum)
  
  # compute the normalization factors
  norm_factor <- mean_sum / sums
  #print(norm_factor)

  # multiply the normalization factors with their corresponding replicate
  MS_Table[, i] <- MS_Table[, i] * norm_factor[1]
  MS_Table[, i + 1] <- MS_Table[, i + 1] * norm_factor[2]
  MS_Table[, i + 2] <- MS_Table[, i + 2] * norm_factor[3]

}
```

We can check whether this worked fine or not.

```{r}
counter <- 0

# there is the need to use round because the numbers seem to be slightly different behind the comma
print(sum(MS_Table[,7]), digits = 17)
print(sum(MS_Table[,8]), digits = 17)

for (i in seq(1, 148, by = 3)) {
  if (round(sum(MS_Table[, i])) != round(sum(MS_Table[, i + 1])) &
      round(sum(MS_Table[, i])) != round(sum(MS_Table[, i + 2]))) {
    counter <- counter + 1
    print(i)
  }
}
counter == 0 # must be TRUE
```

## normalize protein amount per fraction

In the second normalization step, we want to set every protein's amount within one replicate to be 100 per conditions. For example, the protein represented by row 1 has a sum over all 25 fractions of 100 in Ctrl_Rep1. To archive this, we begin with building 6 smaller data frames for every replicate of each condition.

```{r}
# create df that contains all rows of control replicate 1
x <- grep("Ctrl_Rep1", colnames(MS_Table))
Ctrl_Rep1 <- MS_Table[, x]

# create df that contains all rows of RNase replicate 1
y <- grep("RNase_Rep1", colnames(MS_Table))
RNase_Rep1 <- MS_Table[, y]

# create df that contains all rows of control replicate 2
z <- grep("Ctrl_Rep2", colnames(MS_Table))
Ctrl_Rep2 <- MS_Table[, z]

# create df that contains all rows of RNase replicate 2
a <- grep("RNase_Rep2", colnames(MS_Table))
RNase_Rep2 <- MS_Table[, a]

# create df that contains all rows of control replicate 3
b <- grep("Ctrl_Rep3", colnames(MS_Table))
Ctrl_Rep3 <- MS_Table[, b]

# create df that contains all rows of RNase replicate 3
c <- grep("RNase_Rep3", colnames(MS_Table))
RNase_Rep3 <- MS_Table[, c]
```

Following, we can set each row of every replicate's data frame to have a sum of 100.
```{r}
# normalization step 1 of control replicate 1
for (i in 1:nrow(Ctrl_Rep1)) {
  x <- sum(Ctrl_Rep1[i, ]) / 100
  Ctrl_Rep1[i, ] = Ctrl_Rep1[i, ] / x
}

# normalization step 1 of RNase replicate 1
for (i in 1:nrow(RNase_Rep1)) {
  x <- sum(RNase_Rep1[i, ]) / 100
  RNase_Rep1[i, ] = RNase_Rep1[i, ] / x
}

# normalization step 1 of control replicate 2
for (i in 1:nrow(Ctrl_Rep2)) {
  x <- sum(Ctrl_Rep2[i, ]) / 100
  Ctrl_Rep2[i, ] = Ctrl_Rep2[i, ] / x
}

# normalization step 1 of RNase replicate 2
for (i in 1:nrow(RNase_Rep2)) {
    x <- sum(RNase_Rep2[i, ]) / 100
    RNase_Rep2[i, ] = RNase_Rep2[i, ] / x
}

# normalization step 1 of control replicate 3
for (i in 1:nrow(Ctrl_Rep3)) {
  x <- sum(Ctrl_Rep3[i, ]) / 100
  Ctrl_Rep3[i, ] = Ctrl_Rep3[i, ] / x
}

# normalization step 1 of RNase replicate 3
for (i in 1:nrow(RNase_Rep3)) {
  x <- sum(RNase_Rep3[i, ]) / 100
  RNase_Rep3[i, ] = RNase_Rep3[i, ] / x
}
```

Now, we can verify if the sum of every row really equals 100. If everything went correctly, we should get 6 times TRUE!
```{r}
# test if the sum of each row divided by 100 is the number of rows will tell us whether the sum is 100
sum(Ctrl_Rep1) / 100 == nrow(Ctrl_Rep1)
sum(RNase_Rep1) / 100 == nrow(RNase_Rep1)
sum(Ctrl_Rep2) / 100 == nrow(Ctrl_Rep2)
sum(RNase_Rep2) / 100 == nrow(RNase_Rep2)
sum(Ctrl_Rep3) / 100 == nrow(Ctrl_Rep3)
sum(RNase_Rep3) / 100 == nrow(RNase_Rep3)
```

Seems like everything went fine! Now, we can combine the replicate's data frames again to make one single data frame containing all normalized data.

```{r}
df_normalized <- as.data.frame(cbind(Ctrl_Rep1, Ctrl_Rep2, Ctrl_Rep3, RNase_Rep1, RNase_Rep2, RNase_Rep3))
dim(df_normalized) == dim(MS_Table)
sum(df_normalized[1,]) / 6 == 100
df_normalized_copy <- df_normalized
```

## evaluating reproducibility of our data

We want to evaluate the reproducibility of our data by computing the correlation between the replicates of the two conditions for every protein. By doing it that way, we can not only evaluate similarities between the replicates of one condition, we can also sort out single proteins which data do not seem to be reproducible. (First, we create a vector for every protein containing all its values of one replicate, for every replicate and condition.) 

```{r}
tdf_normalized <- as.data.frame(t(df_normalized))

# correlation between ctrl1 and ctrl2 for every protein
c1c2 <- c()
for (i in 1:ncol(tdf_normalized)) {
  c1c2 <- append(c1c2, (cor(tdf_normalized[1:25, i], tdf_normalized[26:50, i])))
}

# correlation between ctrl1 and ctrl3 for every protein
c1c3 <- c()
for (i in 1:ncol(tdf_normalized)) {
  c1c3 <- append(c1c3, (cor(tdf_normalized[1:25, i], tdf_normalized[51:75, i])))
}
 
# correlation between ctrl2 and ctrl3 for every protein 
c2c3 <- c()
for (i in 1:ncol(tdf_normalized)) {
  c2c3 <- append(c2c3, (cor(tdf_normalized[26:50, i], tdf_normalized[51:75, i])))
}

# correlation between rnase1 and rnase2 for every protein
r1r2 <- c()
for (i in 1:ncol(tdf_normalized)) {
  r1r2 <- append(r1r2, (cor(tdf_normalized[76:100, i], tdf_normalized[101:125, i])))
}

# correlation between rnase1 an rnase3 for every protein
r1r3 <- c()
for (i in 1:ncol(tdf_normalized)) {
  r1r3 <- append(r1r3, (cor(tdf_normalized[76:100, i], tdf_normalized[126:150, i])))
}
  
# correlation between rnase2 and rnase3 for every protein
r2r3 <- c()
for (i in 1:ncol(tdf_normalized)) {
  r2r3 <- append(r2r3, (cor(tdf_normalized[101:125, i], tdf_normalized[126:150, i])))
}
```

In order to test which threshold fits the best we try different ones, and plot how many proteins would be removed. We are testing threshold values from 0 to 0.8. Everything above would mean throwing away too many proteins, with high correlation. 

```{r}
tres_values <- c()

for (t in seq(0, 0.8, by = 0.05)) {
  tc <- c()
  
  for (i in 1:ncol(tdf_normalized)) {
    if (c1c2[i] < t) {
      tc <- append(tc, colnames(tdf_normalized)[i])
    }
  }
  for (i in 1:ncol(tdf_normalized)) {
    if (c1c3[i] < t) {
      tc <- append(tc, colnames(tdf_normalized)[i])
    }
  }
  for (i in 1:ncol(tdf_normalized)) {
    if (c2c3[i] < t) {
      tc <- append(tc, colnames(tdf_normalized)[i])
    }
  }
  for (i in 1:ncol(tdf_normalized)) {
    if (r1r2[i] < t) {
      tc <- append(tc, colnames(tdf_normalized)[i])
    }
  }
  for (i in 1:ncol(tdf_normalized)) {
    if (r1r3[i] < t) {
      tc <- append(tc, colnames(tdf_normalized)[i])
    }
  }
  for (i in 1:ncol(tdf_normalized)) {
    if (r2r3[i] < t) {
      tc <- append(tc, colnames(tdf_normalized)[i])
    }
  }
  tres_values <- append(tres_values, length(unique(tc)))
}
```

To evaluate a fitting threshold value we are going to plot the number of proteins which would be removed against the respective threshold value. 

```{r}
plot(seq(0, 0.8, 0.05), tres_values, type = "b", xlab = "Threshold value", ylab = "Number of removed proteins", main = "Evaluating the reproducibility of our data"); abline(v = 0.6, col = "red", lwd = 1, lty = 2)
```
Value 0.6 seems to be a good fit. We can also have a look at the correlation values we just computed by visualizing them with a plot. For example, we can have a look at the distribution:

```{r}
cor_vec <- c(c1c2, c1c3, c2c3, r1r2, r1r3, r2r3)
hist(cor_vec, ylim = c(0, 1000), xlab = "Correlation value", main = "Histogram of correlation values")
```
Here it also looks like 0.6 is a good threshold value. 

To visualize the analysis we just performed, we can plot two replicates which are not strongly correlated and two replicates which are strongly correlated.
```{r}
par(mfrow = c(1, 2))

# no strong correlation
colnames(tdf_normalized)[which(r1r3 == min(r1r3))]
df_plot1 <-
  data.frame(x = tdf_normalized$K2C1_HUMAN[76:100], y = tdf_normalized$K2C1_HUMAN[126:150])
plot1 <- ggplot(df_plot1, aes(x, y)) +
  geom_point() +
  geom_smooth() +
  labs(x = "K2C1_HUMAN RNase Rep1", y = "K2C1_HUMAN RNase Rep3") +
  ggtitle("Low correlation example")

# strong correlation
set.seed(124)
x <- which(c1c2 < 0.97)
colnames(tdf_normalized)[sample(which(c1c2[x] > 0.90), 1)]

df_plot2 <-
  data.frame(x = tdf_normalized$ECE1_HUMAN[1:25] , y = tdf_normalized$ECE1_HUMAN[26:50])
plot2 <- ggplot(df_plot2, aes(x, y)) +
  geom_point() +
  geom_smooth() +
  labs(x = "ECE1_HUMAN Ctrl Rep1", y = "ECE1_HUMAN Ctrl Rep2") +
  ggtitle("High correlation example")

# plot
cowplot::plot_grid(plot1, plot2, labels = "AUTO")
```

Now, we must remove the proteins with correlation values below 0.6.

```{r}
t = 0.6
tc <- c()

for (i in 1:ncol(tdf_normalized)) {
  if (c1c2[i] < t) {
    tc <- append(tc, colnames(tdf_normalized)[i])
  } 
}
for (i in 1:ncol(tdf_normalized)) {
  if (c1c3[i] < t) {
    tc <- append(tc, colnames(tdf_normalized)[i])
  } 
}
for (i in 1:ncol(tdf_normalized)) {
  if (c2c3[i] < t) {
    tc <- append(tc, colnames(tdf_normalized)[i])
  } 
}
for (i in 1:ncol(tdf_normalized)) {
  if (r1r2[i] < t) {
    tc <- append(tc, colnames(tdf_normalized)[i])
  } 
}
for (i in 1:ncol(tdf_normalized)) {
  if (r1r3[i] < t) {
    tc <- append(tc, colnames(tdf_normalized)[i])
  } 
}
for (i in 1:ncol(tdf_normalized)) {
  if (r2r3[i] < t) {
    tc <- append(tc, colnames(tdf_normalized)[i])
  } 
}

length(unique(tc))
utc <- unique(tc)
remove_proteins <- c()

for (i in 1:length(utc)) {
  remove_proteins <- append(remove_proteins, which(rownames(df_normalized) == utc[i]))
}
if (length(remove_proteins) != 0) {
  df_normalized <- df_normalized[-remove_proteins,]
}
dim(df_normalized)
```

Now, we have removed 25 proteins due to an expected low reproducibility. Together with the 7 ones removed in the beginning, we have removed a total of 32 proteines, which is 0.4% of the proteins.
We can save this data frame as a new file **RDeeP_HeLa_Mitosis_normalized.rds**:

```{r}
#saveRDS(df_normalized, file = "RDeeP_HeLa_Mitosis_normalized.rds")
```

```{r}
#df_normalized <- readRDS("RDeeP_HeLa_Mitosis_normalized.rds")
#sum(df_normalized) / 600 == nrow(df_normalized)
```

## combining replicates using mean value

In this step, we want to combine the three replicates of every condition by using the mean value of the three replicates' values. To do so, we first split the `df_normalized` data frame into one data frame containing only the control replicates and one data frame containing only the RNase replicates.

```{r}
df_Ctrl <- df_normalized[, grep("Ctrl", colnames(df_normalized))]
df_RNase <- df_normalized[, grep("RNase", colnames(df_normalized))]
```

Now, we can process the two data frames so that they contain only the mean values of the triplicates.

```{r}
# control condition
df_Ctrl_c <- data.frame()
for (i in 1:nrow(df_Ctrl)) {
  for (j in 1:25) {
    x <- mean(c(df_Ctrl[i, j], df_Ctrl[i, j + 25], df_Ctrl[i, j + 50]))
    df_Ctrl_c[i, j] = x
  }
}

# RNase condition
df_RNase_c <- data.frame()
for (i in 1:nrow(df_RNase)) {
  for (j in 1:25) {
    x <- mean(c(df_RNase[i, j], df_RNase[i, j + 25], df_RNase[i, j + 50]))
    df_RNase_c[i, j] = x
  }
}
```

```{r}
# test if mean values in df_Ctrl_c are computed correctly
score1 <- 0
for (i in 1:nrow(df_Ctrl)) {
  for (j in 1:25) {
    if (mean(c(df_Ctrl[i, j], df_Ctrl[i, j + 25], df_Ctrl[i, j + 50])) != df_Ctrl_c[i, j]) {
      score1 <- score1 + 1
    }
  }
}
score1 == 0

# test if mean values in df_RNase_c are computed correctly
score2 <- 0
for (i in 1:nrow(df_RNase)) {
  for (j in 1:25) {
    if (mean(c(df_RNase[i, j], df_RNase[i, j + 25], df_RNase[i, j + 50])) != df_RNase_c[i, j]) {
      score2 <- score2 + 1
    }
  }
}
score2 == 0
```

Now both, the Ctrl and the RNase data frames, consist of the mean values of the replicates and are scaled to 100 units of each protein. We can now join the both data frames to one again, containing all proteins but only 50 instead of 150 columns.

```{r}
df_normalized_combined <- cbind(df_Ctrl_c, df_RNase_c)
rownames(df_normalized_combined) <- rownames(df_normalized)

cnc <- sapply(1:25, function(x) {paste("Fraction", x, "_Ctrl", sep = "")})
cnr <- sapply(1:25, function(x) {paste("Fraction", x, "_RNase", sep = "")})
colnames(df_normalized_combined)[1:25] <- cnc
colnames(df_normalized_combined)[26:50] <- cnr
```

Lastly, we can save that data frame as a file again to avoid running the code above again and again.

```{r}
# for report
#saveRDS(tres_values, file = "tres_values.rds")

#saveRDS(df_normalized_combined, file = "RDeeP_HeLa_Mitosis_normalized_combined.rds")
```





# PART 2



## load data
 
```{r}
#df_normalized_combined <- readRDS("RDeeP_HeLa_Mitosis_normalized_combined.rds")
```

```{r}
dim(df_normalized_combined)
```

## global maxima for combined values

Here, we want to identify the global maximum for every protein under both conditions using the normalized and combined data.

```{r}
d <- dim(df_normalized_combined)
rows <- d[1] 
columns <- d[2]
fractions_ctrl <- 25
fractions_RNase <- 25

df_global_maxima <- data.frame()
proteins <- rownames(df_normalized_combined)
fractionsnames <- colnames(df_normalized_combined)

for (z in 1:rows) {
  proteinname <- proteins[z]
  
  maxfc <- df_normalized_combined[z, 1]
  maxfcc <- 1
  
  for (fc in 2:25) {
    
    if (maxfc < df_normalized_combined[z, fc]) {
      maxfc <- df_normalized_combined[z, fc]
      maxfcc <- fc
    }
  }
  
  maxrna <- df_normalized_combined[z, 26]
  maxfcr <- 26
  
  for (fc in 27:50) {
    
    if (maxrna < df_normalized_combined[z, fc]) {
      maxrna <- df_normalized_combined[z, fc]
      maxfcr <- fc
    }
  }
  
  df_global_maxima <-
    rbind(
      df_global_maxima,
      data.frame(
        #global_maximum_ctrl_fraction1 = fractionsnames[maxfcc],
        ctrl_global_maximum_fraction = maxfcc,
        ctrl_global_maximum_value = maxfc,
        #global_maximum_rnase_fraction1 = fractionsnames[maxfcr],
        rnase_global_maximum_fraction = maxfcr - 25,
        rnase_global_maximum_value = maxrna
      )
    )
}

rownames(df_global_maxima) <- rownames(df_normalized_combined)
#View(df_global_maxima)
dim(df_global_maxima)
```

```{r}
# test whether both ways work the same, NOT IMPORTANT FOR REPORT
ctrl_global_maximum_fraction <- c()
ctrl_global_maximum_value <- c()
rnase_global_maximum_fraction <- c()
rnase_global_maximum_value <- c()

for (i in 1:nrow(df_normalized_combined)) {
  ctrl_global_maximum_value <-
    append(ctrl_global_maximum_value, max(df_normalized_combined[i, 1:25]))
  ctrl_global_maximum_fraction <-
    append(ctrl_global_maximum_fraction,
           which(df_normalized_combined[i, 1:25] == max(df_normalized_combined[i, 1:25])))
  rnase_global_maximum_value <-
    append(rnase_global_maximum_value, max(df_normalized_combined[i, 26:50]))
  rnase_global_maximum_fraction <-
    append(rnase_global_maximum_fraction,
           which(df_normalized_combined[i, 26:50] == max(df_normalized_combined[i, 26:50])))
  
}

df_global_maxima_test <-
  data.frame(
    row.names = rownames(df_normalized_combined),
    ctrl_global_maximum_fraction = ctrl_global_maximum_fraction,
    ctrl_global_maximum_value = ctrl_global_maximum_value,
    rnase_global_maximum_fraction = rnase_global_maximum_fraction,
    rnase_global_maximum_value = rnase_global_maximum_value
  )

sum(df_global_maxima != df_global_maxima_test)
```

```{r}
#saveRDS(df_global_maxima, file = "RDeeP_HeLa_Mitosis_global_maxima.rds")
```

## global maxima for every replicate

In order to use the function above for every replicate, we need to split the data frame df_normalized into 6 data frames.

```{r}
df_normalized <- readRDS("RDeeP_HeLa_Mitosis_normalized.rds")

df_ctrl1 <- df_normalized[,grep("Ctrl_Rep1", colnames(df_normalized))]
df_ctrl2 <- df_normalized[,grep("Ctrl_Rep2", colnames(df_normalized))]
df_ctrl3 <- df_normalized[,grep("Ctrl_Rep3", colnames(df_normalized))]
df_rnase1 <- df_normalized[,grep("RNase_Rep1", colnames(df_normalized))]
df_rnase2 <- df_normalized[,grep("RNase_Rep2", colnames(df_normalized))]
df_rnase3 <- df_normalized[,grep("RNase_Rep3", colnames(df_normalized))]
```

```{r}
find_global_maxima <- function(df) {
  
  d <- dim(df)
  rows <- d[1]
  columns <- d[2]
  fractions_ctrl <- 25
  #fractions_RNase <- 25
  
  df_output <- data.frame()
  proteins <- rownames(df)
  fractionsnames <- colnames(df)
  
  for (z in 1:rows) {
    proteinname <- proteins[z]
    
    maxfc <- df[z, 1]
    maxfcc <- 1
    
    for (fc in 2:25) {
      if (maxfc < df[z, fc]) {
        maxfc <- df[z, fc]
        maxfcc <- fc
      }
    }
    
    #maxrna <- df[z, 26]
    #maxfcr <- 26
    
    #for (fc in 27:50) {
      #if (maxrna < df[z, fc]) {
        #maxrna <- df[z, fc]
        #maxfcr <- fc
      #}
    #}
    
    df_output <-
      rbind(
        df_output,
        data.frame(
          #global_maximum_ctrl_fraction1 = fractionsnames[maxfcc],
          ctrl_global_maximum_fraction = maxfcc,
          ctrl_global_maximum_value = maxfc
          #global_maximum_rnase_fraction1 = fractionsnames[maxfcr],
          #rnase_global_maximum_fraction = maxfcr - 25,
          #rnase_global_maximum_value = maxrna
        )
      )
  }
  
  rownames(df_output) <- rownames(df)
  #View(df_global_maxima)
  #dim(df_output)
  
  return(df_output)
}
```

```{r}
df_ctrl1_global_maxima <- find_global_maxima(df_ctrl1)
df_ctrl2_global_maxima <- find_global_maxima(df_ctrl2)
df_ctrl3_global_maxima <- find_global_maxima(df_ctrl3)
df_rnase1_global_maxima <- find_global_maxima(df_rnase1)
df_rnase2_global_maxima <- find_global_maxima(df_rnase2)
df_rnase3_global_maxima <- find_global_maxima(df_rnase3)

df_global_maxima_rep <-
  cbind(
    df_ctrl1_global_maxima,
    df_ctrl2_global_maxima,
    df_ctrl3_global_maxima,
    df_rnase1_global_maxima,
    df_rnase2_global_maxima,
    df_rnase3_global_maxima
  )
colnames(df_global_maxima_rep) <-
  c(
    "ctrl1_global_maximum_fraction",
    "ctrl1_global_maximum_value",
    "ctrl2_global_maximum_fraction",
    "ctrl2_global_maximum_value",
    "ctrl3_global_maximum_fraction",
    "ctrl3_global_maximum_value",
    "rnase1_global_maximum_fraction",
    "rnase1_global_maximum_value",
    "rnase2_global_maximum_fraction",
    "rnase2_global_maximum_value",
    "rnase3_global_maximum_fraction",
    "rnase3_global_maximum_value"
  )
```

```{r}
#saveRDS(df_global_maxima_rep, file = "RDeeP_HeLa_Mitosis_global_maxima_rep.rds")
```

## local maxima 

In this part, we also want to identify local maxima for every protein and condition. A local maxima must fulfill three criteria:

- must not be greater than the global maximum
- must be greater than the two neighbouring pairs
- must be greater than 4

### local maxima ctrl

```{r}
# create df that will contain the local maxima, because we do not know yet how many local maxima each protein will have we need to make space for more than one per protein and cut the rest of later

df_local_maxima_ctrl <-
  data.frame(
    row.names = rownames(df_normalized_combined),
    ctrl_local_maximum1_fraction = matrix(0, nrow(df_normalized_combined), 1),
    ctrl_local_maximum1_value = matrix(0, nrow(df_normalized_combined), 1),
    ctrl_local_maximum2_fraction = matrix(0, nrow(df_normalized_combined), 1),
    ctrl_local_maximum2_value = matrix(0, nrow(df_normalized_combined), 1),
    ctrl_local_maximum3_fraction = matrix(0, nrow(df_normalized_combined), 1),
    ctrl_local_maximum3_value = matrix(0, nrow(df_normalized_combined), 1),
    ctrl_local_maximum4_fraction = matrix(0, nrow(df_normalized_combined), 1),
    ctrl_local_maximum4_value = matrix(0, nrow(df_normalized_combined), 1),
    ctrl_local_maximum5_fraction = matrix(0, nrow(df_normalized_combined), 1),
    ctrl_local_maximum5_value = matrix(0, nrow(df_normalized_combined), 1)
  )
```

```{r}
for (i in 1:nrow(df_normalized_combined)) {
  # first loop iterating through the proteins
  over_4 <- c()
  # store indices of values over 4 for every protein in here (only temporary)
  
  for (j in 1:25) {
    # second loop iterating through all Ctrl fractions
    if (df_normalized_combined[i, j] > 4) {
      # only values > 4 are saved
      over_4 <- append(over_4, j)
    }
  }
  #print(over_4)
  local_maxima <- c()
  # store indices of values over 4 and with no higher neighbours (only temporary)
  
  for (v in over_4) {
    # iterate through all values that are > 4
    #print(df_normalized_combined[i, v])
    
    # values on boarders have to be treated differently
    
    # first value (no value to the left)
    if (v == 1) {
      if (df_normalized_combined[i, v] > df_normalized_combined[i, v + 1] &
          df_normalized_combined[i, v] > df_normalized_combined[i, v + 2] &
          df_normalized_combined[i, v] != max(df_normalized_combined[i, 1:25])) {
        local_maxima <- append(local_maxima, v)
      }
      
      # second value (only one value to the left)
    } else if (v == 2) {
      if (df_normalized_combined[i, v] > df_normalized_combined[i, v - 1] &
          df_normalized_combined[i, v] > df_normalized_combined[i, v + 1] &
          df_normalized_combined[i, v] > df_normalized_combined[i, v + 2] &
          df_normalized_combined[i, v] != max(df_normalized_combined[i, 1:25])) {
        local_maxima <- append(local_maxima, v)
      }
      
      # 24th value (only one value to the right)
    } else if (v == 24) {
      if (df_normalized_combined[i, v] > df_normalized_combined[i, v - 1] &
          df_normalized_combined[i, v] > df_normalized_combined[i, v - 2] &
          df_normalized_combined[i, v] > df_normalized_combined[i, v + 1] &
          df_normalized_combined[i, v] != max(df_normalized_combined[i, 1:25])) {
        local_maxima <- append(local_maxima, v)
      }
      
      # 25th value (no value to the right)
    } else if (v == 25) {
      if (df_normalized_combined[i, v] > df_normalized_combined[i, v - 1] &
          df_normalized_combined[i, v] > df_normalized_combined[i, v - 2] &
          df_normalized_combined[i, v] != max(df_normalized_combined[i, 1:25])) {
        local_maxima <- append(local_maxima, v)
      }
      
      # all other values to not have to be treated differently
    } else {
      if (df_normalized_combined[i, v] > df_normalized_combined[i, v - 1] &
          df_normalized_combined[i, v] > df_normalized_combined[i, v - 2] &
          df_normalized_combined[i, v] > df_normalized_combined[i, v + 1] &
          df_normalized_combined[i, v] > df_normalized_combined[i, v + 2] &
          df_normalized_combined[i, v] != max(df_normalized_combined[i, 1:25])) {
        local_maxima <- append(local_maxima, v)
      }
    }
  }
  #print(paste("protein", as.character(i)))
  #print(local_maxima)
  
  # now we need to save the protein name as well as the local maximum's position and value
  
  if (length(local_maxima) == 1) {
    # proteins with one local maximum
    
    df_local_maxima_ctrl[i, 1] <- local_maxima[1]
    df_local_maxima_ctrl[i, 2] <- df_normalized_combined[i, local_maxima[1]]
    
  } else if (length(local_maxima) == 2) {
    # proteins with 2 local maxima
    
    df_local_maxima_ctrl[i, 1] <- local_maxima[1]
    df_local_maxima_ctrl[i, 2] <- df_normalized_combined[i, local_maxima[1]]
    df_local_maxima_ctrl[i, 3] <- local_maxima[2]
    df_local_maxima_ctrl[i, 4] <-df_normalized_combined[i, local_maxima[2]]
    
  } else if (length(local_maxima) == 3) {
    # proteins with 3 local maxima
    
    df_local_maxima_ctrl[i, 1] <- local_maxima[1]
    df_local_maxima_ctrl[i, 2] <- df_normalized_combined[i, local_maxima[1]]
    df_local_maxima_ctrl[i, 3] <- local_maxima[2]
    df_local_maxima_ctrl[i, 4] <-df_normalized_combined[i, local_maxima[2]]
    df_local_maxima_ctrl[i, 5] <- local_maxima[3]
    df_local_maxima_ctrl[i, 6] <-df_normalized_combined[i, local_maxima[3]]
    
  } else if (length(local_maxima) == 4) {
    # proteins that have 4 local maxima
    
    df_local_maxima_ctrl[i, 1] <- local_maxima[1]
    df_local_maxima_ctrl[i, 2] <- df_normalized_combined[i, local_maxima[1]]
    df_local_maxima_ctrl[i, 3] <- local_maxima[2]
    df_local_maxima_ctrl[i, 4] <-df_normalized_combined[i, local_maxima[2]]
    df_local_maxima_ctrl[i, 5] <- local_maxima[3]
    df_local_maxima_ctrl[i, 6] <-df_normalized_combined[i, local_maxima[3]]
    df_local_maxima_ctrl[i, 7] <- local_maxima[4]
    df_local_maxima_ctrl[i, 8] <-df_normalized_combined[i, local_maxima[4]]
    
  } else if (length(local_maxima) == 5) {
    # proteins with 5  local maxima
    
    df_local_maxima_ctrl[i, 1] <- local_maxima[1]
    df_local_maxima_ctrl[i, 2] <- df_normalized_combined[i, local_maxima[1]]
    df_local_maxima_ctrl[i, 3] <- local_maxima[2]
    df_local_maxima_ctrl[i, 4] <-df_normalized_combined[i, local_maxima[2]]
    df_local_maxima_ctrl[i, 5] <- local_maxima[3]
    df_local_maxima_ctrl[i, 6] <-df_normalized_combined[i, local_maxima[3]]
    df_local_maxima_ctrl[i, 7] <- local_maxima[4]
    df_local_maxima_ctrl[i, 8] <-df_normalized_combined[i, local_maxima[4]]
    df_local_maxima_ctrl[i, 9] <- local_maxima[5]
    df_local_maxima_ctrl[i, 10] <-df_normalized_combined[i, local_maxima[5]]
    
  }
}
```

### local maxima rnase

```{r}
# create df that will contain the local maxima, because we do not know yet how many local maxima each protein will have we need to make space for more than one per protein and cut the rest of later

df_local_maxima_rnase <-
  data.frame(
    row.names = rownames(df_normalized_combined),
    rnase_local_maximum1_fraction = matrix(0, nrow(df_normalized_combined), 1),
    rnase_local_maximum1_value = matrix(0, nrow(df_normalized_combined), 1),
    rnase_local_maximum2_fraction = matrix(0, nrow(df_normalized_combined), 1),
    rnase_local_maximum2_value = matrix(0, nrow(df_normalized_combined), 1),
    rnase_local_maximum3_fraction = matrix(0, nrow(df_normalized_combined), 1),
    rnase_local_maximum3_value = matrix(0, nrow(df_normalized_combined), 1),
    rnase_local_maximum4_fraction = matrix(0, nrow(df_normalized_combined), 1),
    rnase_local_maximum4_value = matrix(0, nrow(df_normalized_combined), 1),
    rnase_local_maximum5_fraction = matrix(0, nrow(df_normalized_combined), 1),
    rnase_local_maximum5_value = matrix(0, nrow(df_normalized_combined), 1)
  )
```

```{r}
for (i in 1:nrow(df_normalized_combined)) {
  # first loop iterating through the proteins
  over_4 <- c()
  # store indices of values over 4 for every protein in here (only temporary)
  
  for (j in 26:50) {
    # second loop iterating through all Ctrl fractions
    if (df_normalized_combined[i, j] > 4) {
      # only values > 4 are saved
      over_4 <- append(over_4, j)
    }
  }
  #print(over_4)
  local_maxima <- c()
  # store indices of values over 4 and with no higher neighbours (only temporary)
  
  for (v in over_4) {
    # iterate through all values that are > 4
    #print(df_normalized_combined[i, v])
    
    # values on boarders have to be treated differently
    
    # 26th value (no value to the left)
    if (v == 26) {
      if (df_normalized_combined[i, v] > df_normalized_combined[i, v + 1] &
          df_normalized_combined[i, v] > df_normalized_combined[i, v + 2] &
          df_normalized_combined[i, v] != max(df_normalized_combined[i, 26:50])) {
        local_maxima <- append(local_maxima, v)
      }
      
      # 27th value (only one value to the left)
    } else if (v == 27) {
      if (df_normalized_combined[i, v] > df_normalized_combined[i, v - 1] &
          df_normalized_combined[i, v] > df_normalized_combined[i, v + 1] &
          df_normalized_combined[i, v] > df_normalized_combined[i, v + 2] &
          df_normalized_combined[i, v] != max(df_normalized_combined[i, 26:50])) {
        local_maxima <- append(local_maxima, v)
      }
      
      # 49th value (only one value to the right)
    } else if (v == 49) {
      if (df_normalized_combined[i, v] > df_normalized_combined[i, v - 1] &
          df_normalized_combined[i, v] > df_normalized_combined[i, v - 2] &
          df_normalized_combined[i, v] > df_normalized_combined[i, v + 1] &
          df_normalized_combined[i, v] != max(df_normalized_combined[i, 26:50])) {
        local_maxima <- append(local_maxima, v)
      }
      
      # 50th value (no value to the right)
    } else if (v == 50) {
      if (df_normalized_combined[i, v] > df_normalized_combined[i, v - 1] &
          df_normalized_combined[i, v] > df_normalized_combined[i, v - 2] &
          df_normalized_combined[i, v] != max(df_normalized_combined[i, 26:50])) {
        local_maxima <- append(local_maxima, v)
      }
      
      # all other values to not have to be treated differently
    } else {
      if (df_normalized_combined[i, v] > df_normalized_combined[i, v - 1] &
          df_normalized_combined[i, v] > df_normalized_combined[i, v - 2] &
          df_normalized_combined[i, v] > df_normalized_combined[i, v + 1] &
          df_normalized_combined[i, v] > df_normalized_combined[i, v + 2] &
          df_normalized_combined[i, v] != max(df_normalized_combined[i, 26:50])) {
        local_maxima <- append(local_maxima, v)
      }
    }
  }
  #print(paste("protein", as.character(i)))
  #print(local_maxima)
  
  # now we need to save the protein name as well as the local maximum's position and value
  
  if (length(local_maxima) == 1) {
    # proteins with one local maximum
    
    df_local_maxima_rnase[i, 1] <- local_maxima[1] - 25
    df_local_maxima_rnase[i, 2] <- df_normalized_combined[i, local_maxima[1]]
    
  } else if (length(local_maxima) == 2) {
    # proteins with 2 local maxima
    
    df_local_maxima_rnase[i, 1] <- local_maxima[1] - 25
    df_local_maxima_rnase[i, 2] <- df_normalized_combined[i, local_maxima[1]]
    df_local_maxima_rnase[i, 3] <- local_maxima[2] - 25
    df_local_maxima_rnase[i, 4] <-df_normalized_combined[i, local_maxima[2]]
    
  } else if (length(local_maxima) == 3) {
    # proteins with 3 local maxima
    
    df_local_maxima_rnase[i, 1] <- local_maxima[1] - 25
    df_local_maxima_rnase[i, 2] <- df_normalized_combined[i, local_maxima[1]]
    df_local_maxima_rnase[i, 3] <- local_maxima[2] - 25
    df_local_maxima_rnase[i, 4] <-df_normalized_combined[i, local_maxima[2]]
    df_local_maxima_rnase[i, 5] <- local_maxima[3] - 25
    df_local_maxima_rnase[i, 6] <-df_normalized_combined[i, local_maxima[3]]
    
  } else if (length(local_maxima) == 4) {
    # proteins that have 4 local maxima
    
    df_local_maxima_rnase[i, 1] <- local_maxima[1] - 25
    df_local_maxima_rnase[i, 2] <- df_normalized_combined[i, local_maxima[1]]
    df_local_maxima_rnase[i, 3] <- local_maxima[2] - 25
    df_local_maxima_rnase[i, 4] <-df_normalized_combined[i, local_maxima[2]]
    df_local_maxima_rnase[i, 5] <- local_maxima[3] - 25
    df_local_maxima_rnase[i, 6] <-df_normalized_combined[i, local_maxima[3]]
    df_local_maxima_rnase[i, 7] <- local_maxima[4] - 25
    df_local_maxima_rnase[i, 8] <-df_normalized_combined[i, local_maxima[4]]
    
  } else if (length(local_maxima) == 5) {
    # proteins with 5  local maxima
    
    df_local_maxima_rnase[i, 1] <- local_maxima[1] - 25
    df_local_maxima_rnase[i, 2] <- df_normalized_combined[i, local_maxima[1]]
    df_local_maxima_rnase[i, 3] <- local_maxima[2] - 25
    df_local_maxima_rnase[i, 4] <-df_normalized_combined[i, local_maxima[2]]
    df_local_maxima_rnase[i, 5] <- local_maxima[3] - 25
    df_local_maxima_rnase[i, 6] <-df_normalized_combined[i, local_maxima[3]]
    df_local_maxima_rnase[i, 7] <- local_maxima[4] - 25
    df_local_maxima_rnase[i, 8] <-df_normalized_combined[i, local_maxima[4]]
    df_local_maxima_rnase[i, 9] <- local_maxima[5] - 25
    df_local_maxima_rnase[i, 10] <-df_normalized_combined[i, local_maxima[5]]
    
  }
}
```

### combine and save local maxima

```{r}
df_local_maxima <- as.data.frame(cbind(df_local_maxima_ctrl, df_local_maxima_rnase))
rownames(df_local_maxima) <- rownames(df_normalized_combined)
```

```{r}
#saveRDS(df_local_maxima, file = "RDeeP_HeLa_Mitosis_local_maxima.rds")
```





# PART 3



## load data

```{r}
#df_normalized <- readRDS("RDeeP_HeLa_Mitosis_normalized.rds")
#df_normalized_combined <- readRDS("RDeeP_HeLa_Mitosis_normalized_combined.rds")
#df_global_maxima <- readRDS("RDeeP_HeLa_Mitosis_global_maxima.rds")
#df_global_maxima_rep <- readRDS("RDeeP_HeLa_Mitosis_global_maxima_rep.rds")
```

## shift of global maxima > n fractions

After we found the global maxima for every protein, we can continue to use this information to identify RDeeP in our data. For now, we are only using the global maxima, local maxima are found too and might be used later as well.

The first selection criteria for RDeeP is that the shift of the global maximum between RNase and Ctrl conditions must be greater than n fractions (n could be 1, 2, 3,...). To find out a good value for n, we will do the analysis for multiple values and check the quality of the results. For this analysis step, we are going to use the combined data (only one replicate).

We will start to write a function which gives us the shift in fractions for every protein examined.

```{r}
# create vector that will contain the shift in fractions for every protein
shift_global <- c()

# iterate through every protein and calculate how large the shift is
for (i in 1:nrow(df_global_maxima)) {
  
  x <- df_global_maxima[i, 3] - df_global_maxima[i, 1]
  shift_global <- append(shift_global, x)
  
}

# now we can see the shift for every protein
names(shift_global) <- rownames(df_global_maxima)
```

By doing it that way, negative numbers indicate left shifts (shift to lower fractions), while positive numbers indicate right shifts (shift to higher fractions).

```{r}
sum(shift_global < 0) # left shift
sum(shift_global > 0) # right shift 
sum(shift_global == 0) # no shift
```

As we can see, only relying on the global maxima, 5008 proteins do not shift, 1409 proteins have a left shift and 710 proteins have a right shift. Now, we need to select how great the left or right shift must be to make a protein a RDeeP-candidate.

```{r}
for (n in 0:5) {
  
  x <- length(which(shift_global < -n))
  y <- length(which(shift_global > n))
  z <- length(shift_global) - (x + y)
  
  print(x)
  print(y)
  print(z)
  
  print("")
}
```

As expected, the greater we set n to be, the less shifting proteins we identify. Moreover, the number of proteins with no shift increases less every time we increase n by 1. Therefore, the by far greatest decrease of the number of shifting proteins is given when increasing n from 0 to 1. It is clear that a shift of only one fraction can happen in this experiment without the need of a RDeeP, so setting n to be 0 does not make sense. Setting it to 1, 2 or 3 clearly makes a difference, but a much smaller than when increasing n from 0 to 1. 

Anyway, that does not really help us choosing the right value for n. 

n = 1
```{r}
n = 1 # shift must be greater than n
shift_proteins <- c() # in here, the proteins which are shifting are saved
shift_proteins_index <- c()
counter <- 0

for (i in 1:nrow(df_global_maxima)) {
  
  if (shift_global[i] < -n) {
    shift_proteins <- append(shift_proteins, rownames(df_global_maxima)[i])
    shift_proteins_index <- append(shift_proteins_index, i)
  } else if (shift_global[i] > n) {
    shift_proteins <- append(shift_proteins, rownames(df_global_maxima)[i])
    shift_proteins_index <- append(shift_proteins_index, i)
  } else if (shift_global[i] < -n | shift_global[i] > n) {
    counter <- counter + 1
  } 
}

counter # must be zero
length(shift_proteins) # shift proteins gives us the names of the proteins which shift more than 1 fraction
length(shift_proteins_index) # shift proteins index gives us the indices of the shifting proteins
```

Tested and verified!
```{r}
df_test <- df_global_maxima[-shift_proteins_index, ]
shift_global_test <- c()

for (i in 1:nrow(df_test)) {
  x <- df_test[i, 3] - df_test[i, 1]
  shift_global_test <- append(shift_global_test, x)
}
which(shift_global_test > 1)
which(shift_global_test < -1)
```

## significant difference of protein amount at the global maximum (combined)

The second criteria a protein must meet concerns the protein amount. Next to having a great enough shift, we want RDeePs to differ significantly in protein amount between RNase and Ctrl condition at the global maxima (we are using the Ctrl maxima). To test this aspect, we are going to use a student's t-test. The fraction which we test will be given by the combined data, so that we only have one global maxima for each protein. The t-test, however, will be performed using the normalized but not combined data to have three sample for every protein and condition.
The data frame df_global_maxima will give us the fraction we need. The protein amount is saved in df_normalized.

```{r}
pval <- c()

for (i in 1:nrow(df_global_maxima)) {
  
  fraction <- df_global_maxima[i, 1]
  ctrl_vec <- df_normalized[i, c(fraction, fraction + 25, fraction + 50)]
  rnase_vec <- df_normalized[i, c(fraction + 75, fraction + 100, fraction + 125)]
  
  if (abs(sum(ctrl_vec)) > 299.99 & abs(sum(rnase_vec) > 299.99)) {
    pval <- append(pval, 1)
  } else {
    x <- t.test(ctrl_vec, rnase_vec)
    pval <- append(pval, x$p.value)
  }
}

pval_adjust <- p.adjust(pval, method = "fdr")
pval_low <- which(pval_adjust < 0.05)
sig_dif_proteins <- rownames(df_global_maxima)[pval_low]
sig_dif_proteins_index <- pval_low

length(sig_dif_proteins) # names of proteins with significant t test (combined values)
length(sig_dif_proteins_index) # index of proteins with significant t test (combined values)
```

## significant difference of protein amount at the global maximum (replicates)

```{r}
pval_rep <- c()

for (i in 1:nrow(df_global_maxima_rep)) {
  
  ctrl_vec <-
    c(df_normalized[i, df_global_maxima_rep[i, 1]],
      df_normalized[i, df_global_maxima_rep[i, 3] + 25],
      df_normalized[i, df_global_maxima_rep[i, 5] + 50])
  rnase_vec <-
    c(df_normalized[i, df_global_maxima_rep[i, 1] + 75],
      df_normalized[i, df_global_maxima_rep[i, 3] + 100],
      df_normalized[i, df_global_maxima_rep[i, 5] + 125])
  
  if (abs(sum(ctrl_vec)) > 299.99 & abs(sum(rnase_vec) > 299.99)) {
    pval_rep <- append(pval_rep, 1)
  } else {
    x <- t.test(ctrl_vec, rnase_vec)
    pval_rep <- append(pval_rep, x$p.value)
  }
}

pval_adjust_rep <- p.adjust(pval_rep, method = "fdr")
pval_low_rep <- which(pval_adjust_rep < 0.05)
sig_dif_proteins_rep <- rownames(df_normalized)[pval_low_rep]
sig_dif_proteins_index_rep <- pval_low_rep

length(sig_dif_proteins_rep) # names of proteins with significant t test (replicate values)
length(sig_dif_proteins_index_rep) # index of proteins with significant t test (replicate values)
```

## RNA-dependent proteins

```{r}
# combined values
rdeep <- c(intersect(sig_dif_proteins, shift_proteins))
rdeep_index <- c(intersect(sig_dif_proteins_index, shift_proteins_index))

length(rdeep)
length(rdeep_index)

# replicates
rdeep_rep <- c(intersect(sig_dif_proteins_rep, shift_proteins))
rdeep_index_rep <- c(intersect(sig_dif_proteins_index_rep, shift_proteins_index))

length(rdeep_rep)
length(rdeep_index_rep)
```

```{r}
# test whether everything went fine

sum(rownames(df_normalized)[rdeep_index] != rdeep)
sum(rownames(df_normalized)[rdeep_index_rep] != rdeep_rep)
```

```{r}
df_rdeep <- data.frame(row.names = rdeep_rep, index = rdeep_index_rep)
#saveRDS(df_rdeep, file = "RDeeP_HeLa_Mitosis_rdeep.rds")
```





# PART 4



## Dimension Reduction

Load the normalized and combined data:
```{r}
#df_normalized_combined <- readRDS("RDeeP_HeLa_Mitosis_normalized_combined.rds")
#df_normalized <- readRDS("RDeeP_HeLa_Mitosis_normalized.rds")
```

Perform pca and umap or only umap or only pca on our data:
```{r}
df <- df_normalized_combined # we can use this variable to easily switch between combined and not combined
  
# df_umap1: pca and umap
pca <- prcomp(df)
df_umap1 <- as.data.frame(umap(pca$x))

# df_umap2 : only umap
df_umap2 <- as.data.frame(umap(df))

# df_pca: only pca
df_pca <- as.data.frame(prcomp(df)$x)
```

Plot the results to see whether they differ or not:
```{r}
ggplot(df_umap1, aes(x = V1, y = V2)) +
  geom_point() +
  ggtitle("pca and umap")

ggplot(df_umap2, aes(x = V1, y = V2)) +
  geom_point() +
  ggtitle("umap")

ggplot(df_pca, aes(x = PC1, y = PC2)) +
  geom_point() + 
  ggtitle("pca")
```

Now, we can colour proteins according to their shifting behaviour.
```{r}
n = 1
df_umap <- cbind(df_umap1, matrix(0, nrow(df_normalized_combined), 1))
colnames(df_umap) <- c("V1", "V2", "shift")

for (i in 1:nrow(df_umap)) {
  if (shift_global[i] < -n) {
    df_umap[i, 3] <- "left shift"
  } else if (shift_global[i] > n) {
    df_umap[i, 3] <- "right shift"
  } else {
    df_umap[i, 3] <- "no shift"
  }
}
```

We can also colour proteins according to our results (non-RDeep or RDeeP).
```{r}
for (i in 1:nrow(df_umap)) {
  if (rownames(df_umap)[i] %in% rownames(df_rdeep)) {
    df_umap$rdeep[i] <- "rdeep"
  } else {
    df_umap$rdeep[i] <- "non rdeep"
  }
}
```

```{r}
alpha <- ifelse(df_umap$shift == "no shift", 0.1, 1)

ggplot(df_umap, aes(x = V1, y = V2, color = shift)) +
  geom_point(alpha = alpha) +
  ggtitle("UMAP of normalized and combined data")
```

```{r}
alpha <- ifelse(df_umap$rdeep == "non rdeep", 0.1, 1)

ggplot(df_umap, aes(x = V1, y = V2, color = rdeep)) +
  geom_point(alpha = alpha) +
  ggtitle("UMAP of normalized and combined data")
```

We can do the same for pca.
```{r}
n = 1
df_pca_shift <- cbind(df_pca, matrix(0, nrow(df_normalized_combined), 1))
colnames(df_pca_shift)[51] <- "shift"

for (i in 1:nrow(df_pca_shift)) {
  if (shift_global[i] < -n) {
    df_pca_shift[i, 51] <- "left shift"
  } else if (shift_global[i] > n) {
    df_pca_shift[i, 51] <- "right shift"
  } else {
    df_pca_shift[i, 51] <- "no shift"
  }
  
}
```

```{r}
alpha <- ifelse(df_pca_shift$shift == "no shift", 0.3, 1)

ggplot(df_pca_shift, aes(x = PC1, y = PC2, color = shift)) +
  geom_point(alpha = alpha) +
  ggtitle("PCA of normalized and combined data")
```

## group proteins according to their shifting behaviour

```{r}
#df_global_maxima <- readRDS("RDeeP_HeLa_Mitosis_global_maxima.rds")
```

```{r}
n = 1
df_global_maxima_shift <- cbind(df_global_maxima, matrix(0, nrow(df_normalized_combined), 1))
colnames(df_global_maxima_shift)[5] <- "shift"

for (i in 1:nrow(df_global_maxima_shift)) {
  if (shift_global[i] < -n) {
    df_global_maxima_shift[i, 5] <- "left shift"
  } else if (shift_global[i] > n) {
    df_global_maxima_shift[i, 5] <- "right shift"
  } else {
    df_global_maxima_shift[i, 5] <- "no shift"
  }
  
}
```

```{r}
ggplot(
  df_global_maxima_shift,
  aes(y = ctrl_global_maximum_fraction, x = rnase_global_maximum_fraction, color = shift)) +
  geom_count() +
  ggtitle("Shifting behaviour of proteins") +
  xlab("global maximum RNase condition") + ylab("global maximum Ctrl conditions")
```

## kmeans with shift

We are going to apply kmeans using the shift of global maxima for every protein.
```{r}
df_cluster <- as.data.frame(cbind(shift_global, pval_adjust_rep))
rownames(df_cluster) <- rownames(df_normalized)

for (i in 1:nrow(df_cluster)) {
  if (rownames(df_cluster)[i] %in% rownames(df_rdeep)) {
    df_cluster$rdeep[i] <- "rdeep"
  } else {
    df_cluster$rdeep[i] <- "non rdeep"
  }
}

df_cluster$ctrl_maximum <- df_global_maxima$ctrl_global_maximum_fraction
df_cluster$rnase_maximum <- df_global_maxima$rnase_global_maximum_fraction
```

First, we want to check how many centers we should define.
```{r}
wss = sapply(2:10, function(x) { 
  kmeans(df_cluster[,1], centers = x, nstart = 100)$tot.withinss
})
plot(2:10, wss, xlab = "numbers of centers", ylab = "total within-cluster sum of squares", type = "b")
```

Due to no kink being visible, we are going to use 3 centers to cluster for no shift, right shift and left shift.
```{r}
km2 <- kmeans(df_cluster[,1], centers = 3)
```

```{r}
df_umap$cluster2 <- as.character(km2$cluster)

ggplot(df_umap, aes(x = V1, y = V2, color = cluster2)) +
  geom_point() +
  ggtitle("UMAP of normalized and combined data, proteins coloured according to kmeans")
```

The results look a lot like the plot above, where we coloured proteins according to their shifting behaviour.





# PART 5



## load data

```{r}
#df_normalized <- readRDS("RDeeP_HeLa_Mitosis_normalized.rds")
#df_normalized_combined <- readRDS("RDeeP_HeLa_Mitosis_normalized_combined.rds")
```

## create data frame for linear regression

```{r}
tdf <- as.data.frame(t(df_normalized_combined))
```

First, we want to create a data frame which hold all important information we need for our regression analysis.
```{r}
# create df that will contain important information for linear modelling
df_lm <- data.frame(row.names = rownames(df_normalized_combined))

# add vector containing correlation between rnase and control condition to df
vec_cor <- apply(tdf, 2, function(x) {
  cor(x[1:25], x[26:50])
})
df_lm$correlation <- vec_cor
df_lm$shift <- shift_global

# add row with absolute shifts
df_lm$ab_shift <- abs(df_lm$shift)

# add differences between rnase and ctrl conditions
for (i in 1:nrow(df_normalized_combined)) {
  
  s <- 0
  
  for (j in 1:25) {
    s <- s + abs(df_normalized_combined[i, j] - df_normalized_combined[i, j + 25])
  }
  
  df_lm$dif[i] <- s
  
}

# add whether every protein is rdeep or not
df_lm$rdeep <- df_umap$rdeep
for (i in 1:nrow(df_lm)) {
  if (df_lm$rdeep[i] == "rdeep") {
    df_lm$rdeep2[i] <- 1
  } else {
    df_lm$rdeep2[i] <- 0
  }
}
```

## predict shift with correlation

RDeePs should have a lower correlation between RNase und Ctrl conditions than non-RDeePs:
```{r}
cor(tdf$EED_HUMAN[1:25], tdf$EED_HUMAN[26:50])  # RDeeP
cor(tdf$`1433B_HUMAN`[1:25], tdf$`1433B_HUMAN`[26:50]) # non-RDeeP
```

Using this, we can try to make a linear regression model. To train and test our model with our data, we split it so that 80% of the will train the model and 20% will test it. The split is random.
```{r}
set.seed(123)
n <- nrow(df_lm)
x <- sample(c(TRUE, FALSE), n, replace=TRUE, prob=c(0.8, 0.2))
df_lm_train <- df_lm[x, ]
df_lm_test <- df_lm[!x, ]
```

Now, we can train the model using df_lm_train.
```{r}
linear_model <- lm(shift ~ correlation, data = df_lm_train)
summary(linear_model)
```

We can now check some things to decide whether this model is valid or not (residuals normally distributed, residuals not correlated to explanatory variable).
```{r}
# normal distribution of residuals? 
hist(linear_model$residuals, breaks = 20, main = "Histogram of residuals", xlab = "Residuals")

qqnorm(linear_model$residuals)
qqline(linear_model$residuals)

## correlation residuals x-values? 
cor(df_lm_train$correlation, linear_model$residuals)
plot(df_lm_train$correlation, linear_model$residuals, pch = 20, main = "Visualizing correlation", ylab = "Residuals", xlab = "explanatory varaible")
```
The residuals seem to be quite normally distributed with mean = 0. They also seem to be not correlated at all with the explanatory variable (correlation between Ctrl and RNase conditions).

After we have trained our model and verified its accuracy, we can test it by using df_lm_test.
```{r}
pm <- predict.lm(linear_model, newdata = df_lm_test, se.fit = TRUE, interval = "confidence")

df_lm_predict <- as.data.frame(cbind(df_lm_test$shift, pm$fit))
```

```{r}
ggplot(df_lm_predict, aes(x = V1, y = fit)) +
  geom_point() +
  geom_abline(col = "blue", lwd = 2) +
  ggtitle("Real values vs. predicted values") +
  xlab("Real values") + ylab("Predicted values")
```

!!! Model describes left shifts well, but right shifts not !!!

We can also try to evaluate how accurate our model is by having a look at the confidence intervals.
```{r}
counter <- 0

for (i in 1:nrow(df_lm_predict)) {
  if (df_lm_predict$V1[i] > df_lm_predict$lwr[i] & df_lm_predict$V1[i] < df_lm_predict$upr[i]) {
    counter <- counter + 1
  }
}

counter / nrow(df_lm_predict) * 100  
```

Only 6% of the real shifts are in between the boarders of the confidence interval...
We can also have a look at the standard errors for every prediction.
```{r}
mean(pm$se.fit)
```

## same model but absolut shift values

```{r}
# df for model training
df_lm_train_ab <- df_lm_train
df_lm_train_ab$shift <- abs(df_lm_train_ab$shift)

# df for model testing
df_lm_test_ab <- df_lm_test
df_lm_test_ab$shift <- abs(df_lm_test_ab$shift)
```

```{r}
linear_model_ab <- lm(shift ~ correlation, data = df_lm_train_ab)
summary(linear_model_ab)
```

```{r}
# normal distribution of residuals? 
hist(linear_model_ab$residuals, breaks = 20, main = "Histogram of residuals", xlab = "Residuals")

qqnorm(linear_model_ab$residuals)
qqline(linear_model_ab$residuals)

## correlation residuals x-values? 
cor(df_lm_train_ab$correlation, linear_model_ab$residuals)
plot(df_lm_train_ab$correlation, linear_model_ab$residuals, pch = 20, main = "Visualizing correlation", ylab = "Residuals", xlab = "explanatory varaible")
```

```{r}
pm_ab <- predict.lm(linear_model_ab, newdata = df_lm_test_ab, se.fit = TRUE, interval = "confidence")

df_lm_predict_ab <- as.data.frame(cbind(df_lm_test_ab$shift, pm_ab$fit))
```

```{r}
ggplot(df_lm_predict_ab, aes(x = V1, y = fit)) +
  geom_point() +
  geom_abline(col = "blue", lwd = 2) +
  ggtitle("Real values vs. predicted values") +
  xlab("Real values") + ylab("Predicted values")
```

```{r}
counter <- 0

for (i in 1:nrow(df_lm_predict_ab)) {
  if (df_lm_predict_ab$V1[i] > df_lm_predict_ab$lwr[i] & df_lm_predict_ab$V1[i] < df_lm_predict_ab$upr[i]) {
    counter <- counter + 1
  }
}

counter / nrow(df_lm_predict_ab) * 100  
```

```{r}
mean(pm_ab$se.fit)
```

## more regression models

regression model: using differences between ctrl and rnase conditions

```{r}
# train linear model
linear_model <- lm(ab_shift ~ correlation + dif, data = df_lm_train)
summary(linear_model)
```

```{r}
# normal distribution of residuals? 
hist(linear_model$residuals, breaks = 20)

qqnorm(linear_model$residuals)
qqline(linear_model$residuals)

## correlation residuals x-values? 
cor(df_lm_train$dif, linear_model$residuals)
plot(df_lm_train$dif, linear_model$residuals, pch = 20)
```

```{r}
pm <- predict.lm(linear_model, newdata = df_lm_test, se.fit = TRUE, interval = "confidence")

df_lm_predict <- as.data.frame(cbind(df_lm_test$ab_shift, pm$fit))
```

```{r}
plot(
  df_lm_predict$V1,
  df_lm_predict$fit,
  pch = 20,
  col = 'blue',
  xlab = 'Real values',
  ylab = 'Predicted values'
); abline(0, 1, col = "blue", lwd = 5)
```

## comparison with data bases

Finally, we want to compare our results with given results (e.g. databases)
```{r}
df_rdeep <- readRDS("RDeeP_HeLa_Mitosis_rdeep.rds")

rdeep <- rownames(df_rdeep)
non_rdeep <- rownames(df_normalized)[-df_rdeep$index]
```

```{r}
load("HS_non_RBPs.RData")
load("HS_RBPs.RData")
```


```{r}
# number of our RDeePs which are RBPs
counter <- 0

for (i in rdeep) {
  if (i %in% HS_RBPs$Entry_Name.x) {
    counter <- counter + 1
  }
}
counter
```

```{r}
# number of our RDeePs which are non RBPs

counter <- 0

for (i in rdeep) {
  if (i %in% HS_non_RBPs$Entry_Name.x) {
    counter <- counter + 1
  }
}
counter
```

```{r}
# number of our non RDeePs which are RBPs

counter <- 0

for (i in non_rdeep) {
  if (i %in% HS_RBPs$Entry_Name.x) {
    counter <- counter + 1
  }
}
counter
```

```{r}
# number of our non RDeePs which are non RBPs

counter <- 0

for (i in non_rdeep) {
  if (i %in% HS_non_RBPs$Entry_Name.x) {
    counter <- counter + 1
  }
}
counter
```

## comparison with local maxima

To have a look at the local maxima as well, we are going to calculate a global profile for every protein's local maxima by computing the squared sums of the maxima's fractions for both conditions. Later, we can use this information to compare the results with our RDeePs in order to see whether leaving out the local maxima was making our results less accurate.

```{r}
df_local_maxima <- readRDS("RDeeP_HeLa_Mitosis_local_maxima.rds")
```

```{r}
# compute squared sums of fractions for both conditions

df_local_maxima_qs <-
  data.frame(row.names = rownames(df_local_maxima))

for (i in 1:nrow(df_local_maxima_qs)) {
  df_local_maxima_qs$ctrl[i] <-
    sqrt(
      df_local_maxima$ctrl_local_maximum1_fraction[i] ** 2 + df_local_maxima$ctrl_local_maximum2_fraction[i] ** 2 +   df_local_maxima$ctrl_local_maximum3_fraction[i] ** 2 + df_local_maxima$ctrl_local_maximum4_fraction[i] ** 2 +     df_local_maxima$ctrl_local_maximum5_fraction[i] ** 2
    )
  
  df_local_maxima_qs$rnase[i] <-
    sqrt(
      df_local_maxima$rnase_local_maximum1_fraction[i] ** 2 + df_local_maxima$rnase_local_maximum2_fraction[i] ** 2 +   df_local_maxima$rnase_local_maximum3_fraction[i] ** 2 + df_local_maxima$rnase_local_maximum4_fraction[i] ** 2 +   df_local_maxima$rnase_local_maximum5_fraction[i] ** 2
    )
}
```

```{r}
# add difference to data frame (absolute difference)

df_local_maxima_qs$dif <- abs(df_local_maxima_qs$ctrl - df_local_maxima_qs$rnase)
shift_proteins_l <- c(rownames(df_local_maxima_qs)[which(df_local_maxima_qs$dif > 5)])
length(shift_proteins_l)
```

```{r}
# RDeePs if global and local

shift_proteins_lg <- c(intersect(shift_proteins, shift_proteins_l))
length(c(intersect(shift_proteins_lg, sig_dif_proteins_rep)))
```

```{r}
# RDeePs if global or local

x1 <- c(intersect(shift_proteins, sig_dif_proteins_rep))
x2 <- c(intersect(shift_proteins_l, sig_dif_proteins_rep))

x3 <- c(x1, x2)
length(unique(x3))
```
