---
title: "Analysis Part 1"
author: "Michel, Lennart, Michèle, Sebastian"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
library("tidyverse")
```

## load data

The data frame is loaded as describe on the Github page for project 3.
```{r}
MS_Table <- read.table("RDeeP_HeLa_Mitosis.csv", header = TRUE, row.names = 1, sep = ";")
```

```{r}
head(rownames(MS_Table),12)
```

```{r}
head(colnames(MS_Table),12)
```

## data cleanup

First, we want to check whether every column's data type is numeric.
```{r}
sum(apply(MS_Table, 2, is.numeric)) == ncol(MS_Table)
glimpse(MS_Table)
```

Every column's data type is numeric (double)! 

We want to remove every row that has only zeros or negative numbers in at least one of the 3 replicates for each condition (RNase or Ctrl). Those rows will not help us to identify shifts and therefore to identify RNA-depending proteins.
```{r}
del_row <- c()
for (i in c("Ctrl_Rep1", "Ctrl_Rep2", "Ctrl_Rep3", "RNase_Rep1", "RNase_Rep2", "RNase_Rep3")) {
  for (j in 1:nrow(MS_Table)) {
    if (sum(MS_Table[j, grep(i, colnames(MS_Table))]) <= 0) {
      del_row <- append(del_row, j)
    }
  }
}
del_row_sort <- sort(unique(del_row))
del_row_sort
```

Therefore, we remove rows 691, 3045, 3575, 4057, 5568, 6732 and 7112.
```{r}
MS_Table <- MS_Table[-del_row_sort,]
```

All in all, we removed 7 rows that contained only zeros in at least one of the replicates of the two conditions (RNase or Control).

## describing the structure of our data

Dimension of our data frame:
```{r}
dim(MS_Table)
```

Some rownames:
```{r}
head(rownames(MS_Table))
```


Some columnnames:
```{r}
head(colnames(MS_Table))
```

## normalize protein amount per fraction

In the first normalization step, we want to set every protein's amount within one replicate to be 100 per conditions. For example, the protein represented by row 1 has a sum over all 25 fractions of 100 in Ctrl_Rep1. To archive this, we begin with building 6 smaller data frames for every replicate of each condition.
```{r}
# create df that contains all rows of control replicate 1
x <- grep("Ctrl_Rep1", colnames(MS_Table))
Ctrl_Rep1 <- MS_Table[, x]

# create df that contains all rows of RNase replicate 1
y <- grep("RNase_Rep1", colnames(MS_Table))
RNase_Rep1 <- MS_Table[, y]

# create df that contains all rows of control replicate 2
z <- grep("Ctrl_Rep2", colnames(MS_Table))
Ctrl_Rep2 <- MS_Table[, z]

# create df that contains all rows of RNase replicate 2
a <- grep("RNase_Rep2", colnames(MS_Table))
RNase_Rep2 <- MS_Table[, a]

# create df that contains all rows of control replicate 3
b <- grep("Ctrl_Rep3", colnames(MS_Table))
Ctrl_Rep3 <- MS_Table[, b]

# create df that contains all rows of RNase replicate 3
c <- grep("RNase_Rep3", colnames(MS_Table))
RNase_Rep3 <- MS_Table[, c]
```

Following, we can set each row of every replicate's data frame to have a sum of 100.
```{r}
# normalization step 1 of control replicate 1
for (i in 1:nrow(Ctrl_Rep1)) {
  x <- sum(Ctrl_Rep1[i, ]) / 100
  Ctrl_Rep1[i, ] = Ctrl_Rep1[i, ] / x
}

# normalization step 1 of RNase replicate 1
for (i in 1:nrow(RNase_Rep1)) {
  x <- sum(RNase_Rep1[i, ]) / 100
  RNase_Rep1[i, ] = RNase_Rep1[i, ] / x
}

# normalization step 1 of control replicate 2
for (i in 1:nrow(Ctrl_Rep2)) {
  x <- sum(Ctrl_Rep2[i, ]) / 100
  Ctrl_Rep2[i, ] = Ctrl_Rep2[i, ] / x
}

# normalization step 1 of RNase replicate 2
for (i in 1:nrow(RNase_Rep2)) {
    x <- sum(RNase_Rep2[i, ]) / 100
    RNase_Rep2[i, ] = RNase_Rep2[i, ] / x
}

# normalization step 1 of control replicate 3
for (i in 1:nrow(Ctrl_Rep3)) {
  x <- sum(Ctrl_Rep3[i, ]) / 100
  Ctrl_Rep3[i, ] = Ctrl_Rep3[i, ] / x
}

# normalization step 1 of RNase replicate 3
for (i in 1:nrow(RNase_Rep3)) {
  x <- sum(RNase_Rep3[i, ]) / 100
  RNase_Rep3[i, ] = RNase_Rep3[i, ] / x
}
```

Now, we can verify if the sum of every row really equals 100. If everything went correctly, we should get 6 times TRUE!
```{r}
sum(Ctrl_Rep1) / 100 == nrow(Ctrl_Rep1)
sum(RNase_Rep1) / 100 == nrow(RNase_Rep1)
sum(Ctrl_Rep2) / 100 == nrow(Ctrl_Rep2)
sum(RNase_Rep2) / 100 == nrow(RNase_Rep2)
sum(Ctrl_Rep3) / 100 == nrow(Ctrl_Rep3)
sum(RNase_Rep3) / 100 == nrow(RNase_Rep3)
```

Seems like everything went fine! Now, we can combine the replicate's data frames again to make one single data frame containing all normalized data.

```{r}
df_normalized <- as.data.frame(cbind(Ctrl_Rep1, Ctrl_Rep2, Ctrl_Rep3, RNase_Rep1, RNase_Rep2, RNase_Rep3))
dim(df_normalized) == dim(MS_Table)
sum(df_normalized[1,]) / 6 == 100
```

We can save this data frame as a new file **RDeeP_HeLa_Mitosis_normalized.rds**:

```{r}
saveRDS(df_normalized, file = "DeeP_HeLa_Mitosis_normalized.rds")
```

```{r}
df_normalized <- readRDS("DeeP_HeLa_Mitosis_normalized.rds")
sum(df_normalized[1,]) / 6 == 100
```

## evaluating reproducibility of our data

We want to evaluate the reproducibility of our data by computing the correlation between the replicates of the two conditions.

## combining replicates using mean value

In this step, we want to combine the three replicates of every condition by using the mean value of the three replicates' values. To do so, we first split the `df_normalized` data frame into one data frame containing only the control replicates and one data frame containing only the RNase replicates.

```{r}
df_Ctrl <- df_normalized[, grep("Ctrl", colnames(df_normalized))]
df_RNase <- df_normalized[, grep("RNase", colnames(df_normalized))]
```

Now, we can process the two data frames so that they contain only the mean values of the triplicates.

```{r}
# control condition
df_Ctrl_c <- data.frame()
for (i in 1:nrow(df_Ctrl)) {
  for (j in 1:25) {
    x <- mean(c(df_Ctrl[i, j], df_Ctrl[i, j + 25], df_Ctrl[i, j + 50]))
    df_Ctrl_c[i, j] = x
  }
}

# RNase condition
df_RNase_c <- data.frame()
for (i in 1:nrow(df_RNase)) {
  for (j in 1:25) {
    x <- mean(c(df_RNase[i, j], df_RNase[i, j + 25], df_RNase[i, j + 50]))
    df_RNase_c[i, j] = x
  }
}
```

```{r}
# test if mean values in df_Ctrl_c are computed correctly
score1 <- 0
for (i in 1:nrow(df_Ctrl)) {
  for (j in 1:25) {
    if (mean(c(df_Ctrl[i, j], df_Ctrl[i, j + 25], df_Ctrl[i, j + 50])) != df_Ctrl_c[i, j]) {
      score1 <- score1 + 1
    }
  }
}
score1 == 0

# test if mean values in df_RNase_c are computed correctly
score2 <- 0
for (i in 1:nrow(df_RNase)) {
  for (j in 1:25) {
    if (mean(c(df_RNase[i, j], df_RNase[i, j + 25], df_RNase[i, j + 50])) != df_RNase_c[i, j]) {
      score2 <- score2 + 1
    }
  }
}
score2 == 0
```

Now both, the Ctrl and the RNase data frames, consist of the mean values of the replicates and are scaled to 100 units of each protein. We can now join the both data frames to one again, containing all proteins but only 50 instead of 150 columns.

```{r}
df_normalized_combined <- cbind(df_Ctrl_c, df_RNase_c)
rownames(df_normalized_combined) <- rownames(MS_Table)

cnc <- sapply(1:25, function(x) {paste("Fraction", x, "_Ctrl", sep = "")})
cnr <- sapply(1:25, function(x) {paste("Fraction", x, "_RNase", sep = "")})
colnames(df_normalized_combined)[1:25] <- cnc
colnames(df_normalized_combined)[26:50] <- cnr
```

Lastly, we can save that data frame as a file again to avoid running the code above again and again.

```{r}
saveRDS(df_normalized_combined, file = "DeeP_HeLa_Mitosis_normalized_combined.rds")
```








