---
title: 'Data Analysis Project 2022: Topic 03 Team 2 - Proteome-wide Screen for RNA-dependent
  Proteins'
author: "Michel Tarnow, Lennart Müller, Michèle Bennek, Sebastian Rickert"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
library("tidyverse")
```

# Data Analysis Project 3.2

## load data

The data frame is loaded as describe on the Github page for project 3.
```{r}
MS_Table <- read.table("RDeeP_HeLa_Mitosis.csv", header = TRUE, row.names = 1, sep = ";")
```

```{r}
head(rownames(MS_Table),12)
```

```{r}
head(colnames(MS_Table),12)
```

## data cleanup

First, we want to check whether every column's data type is numeric.
```{r}
sum(apply(MS_Table, 2, is.numeric)) == ncol(MS_Table)
glimpse(MS_Table)
```

Every column's data type is numeric (double).

We want to remove every row that has only zeros or negative numbers in at least one of the 3 replicates for each condition (RNase or Ctrl). Those rows will not help us to identify shifts and therefore to identify RNA-depending proteins.
```{r}
del_row <- c()
for (i in c("Ctrl_Rep1", "Ctrl_Rep2", "Ctrl_Rep3", "RNase_Rep1", "RNase_Rep2", "RNase_Rep3")) {
  for (j in 1:nrow(MS_Table)) {
    if (sum(MS_Table[j, grep(i, colnames(MS_Table))]) <= 0) {
      del_row <- append(del_row, j)
    }
  }
}
del_row_sort <- sort(unique(del_row))
del_row_sort
```

Therefore, we remove rows 691, 3045, 3575, 4057, 5568, 6732 and 7112.
```{r}
MS_Table <- MS_Table[-del_row_sort,]
```

All in all, we removed 7 rows that contained only zeros in at least one of the replicates of the two conditions (RNase or Control).

## describing data, reproducibilty, normalization

### decribing data

### normalization step 1

In the first normalization step, we want to set every protein's amount within one replicate to be 100 per conditions. For example, the protein represented by row 1 has a sum over all 25 fractions of 100 in Ctrl_Rep1. To archive this, we begin with building 6 smaller data frames for every replicate of each condition.
```{r}
# create df that contains all rows of control replicate 1
x <- grep("Ctrl_Rep1", colnames(MS_Table))
Ctrl_Rep1 <- MS_Table[, x]

# create df that contains all rows of RNase replicate 1
y <- grep("RNase_Rep1", colnames(MS_Table))
RNase_Rep1 <- MS_Table[, y]

# create df that contains all rows of control replicate 2
z <- grep("Ctrl_Rep2", colnames(MS_Table))
Ctrl_Rep2 <- MS_Table[, z]

# create df that contains all rows of RNase replicate 2
a <- grep("RNase_Rep2", colnames(MS_Table))
RNase_Rep2 <- MS_Table[, a]

# create df that contains all rows of control replicate 3
b <- grep("Ctrl_Rep3", colnames(MS_Table))
Ctrl_Rep3 <- MS_Table[, b]

# create df that contains all rows of RNase replicate 3
c <- grep("RNase_Rep3", colnames(MS_Table))
RNase_Rep3 <- MS_Table[, c]
```

Following, we can set each row of every replicate's data frame to have a sum of 100.
```{r}
# normalization step 1 of control replicate 1
for (i in 1:nrow(Ctrl_Rep1)) {
  x <- sum(Ctrl_Rep1[i, ]) / 100
  Ctrl_Rep1[i, ] = Ctrl_Rep1[i, ] / x
}

# normalization step 1 of RNase replicate 1
for (i in 1:nrow(RNase_Rep1)) {
  x <- sum(RNase_Rep1[i, ]) / 100
  RNase_Rep1[i, ] = RNase_Rep1[i, ] / x
}

# normalization step 1 of control replicate 2
for (i in 1:nrow(Ctrl_Rep2)) {
  x <- sum(Ctrl_Rep2[i, ]) / 100
  Ctrl_Rep2[i, ] = Ctrl_Rep2[i, ] / x
}

# normalization step 1 of RNase replicate 2
for (i in 1:nrow(RNase_Rep2)) {
    x <- sum(RNase_Rep2[i, ]) / 100
    RNase_Rep2[i, ] = RNase_Rep2[i, ] / x
}

# normalization step 1 of control replicate 3
for (i in 1:nrow(Ctrl_Rep3)) {
  x <- sum(Ctrl_Rep3[i, ]) / 100
  Ctrl_Rep3[i, ] = Ctrl_Rep3[i, ] / x
}

# normalization step 1 of RNase replicate 3
for (i in 1:nrow(RNase_Rep3)) {
  x <- sum(RNase_Rep3[i, ]) / 100
  RNase_Rep3[i, ] = RNase_Rep3[i, ] / x
}
```

Now, we can verify if the sum of every row really equals 100. If everything went correctly, we should get 6 times TRUE!
```{r}
sum(Ctrl_Rep1) / 100 == nrow(Ctrl_Rep1)
sum(RNase_Rep1) / 100 == nrow(RNase_Rep1)
sum(Ctrl_Rep2) / 100 == nrow(Ctrl_Rep2)
sum(RNase_Rep2) / 100 == nrow(RNase_Rep2)
sum(Ctrl_Rep3) / 100 == nrow(Ctrl_Rep3)
sum(RNase_Rep3) / 100 == nrow(RNase_Rep3)
```

### plotting distribution of proteins

```{r}
par(mfrow = c(1, 2))
plot(
  1:25,
  Ctrl_Rep1[5353, ],
  type = "l",
  main = rownames(Ctrl_Rep1)[5353],
  xlab = "Fraction",
  ylab = "protein amount"
)
abline(v = which(Ctrl_Rep1[5353, ] == max(Ctrl_Rep1[5353, ])), col = "blue")
abline(v = which(RNase_Rep1[5353, ] == max(RNase_Rep1[5353, ])), col = "red")

plot(
  1:25,
  RNase_Rep1[5353, ],
  type = "l",
  main = rownames(RNase_Rep1)[5353],
  xlab = "Fraction",
  ylab = "protein amount"
)
abline(v = which(Ctrl_Rep1[5353, ] == max(Ctrl_Rep1[5353, ])), col = "blue")
abline(v = which(RNase_Rep1[5353, ] == max(RNase_Rep1[5353, ])), col = "red")
```
